### 2016-02-29 abstract.tex
Add abstract.tex

\documentclass[a4paper]{jpconf}
\usepackage{graphicx}
\usepackage{comment}
\usepackage{cite}
\usepackage{gensymb}

\begin{document}

\title{Validation of an Actuator Line Model Coupled to a Dynamic Stall Model for Pitching Motion Characteristic to Vertical Axis Turbine} % \\ \jpcs}
\author{Victor Mendoza$^{1}$, Pete Bachant$^{2}$, Martin Wosnik$^{2}$ and Anders Goude$^{1}$ }
\address{$^{1}$ Department of Engineering Sciences, Division of Electricity, Uppsala University, \\Uppsala 751 21, Sweden}
\address{$^{2}$ Center for Ocean Renewable Energy, University of New Hampshire, 24 Colovos Rd.,\\ Durham, NH 03824, USA}
\ead{victor.mendoza@angstrom.uu.se}

\begin{abstract}
Vertical Axis Turbines (VAWT) have been used for extracting of energy either from sea or wind. A simpler design, low cost of maintenance and the capability of working independent from the direction of the flow are some of the most important advantages over conventional Horizontal Axis Turbines (HAWTs). Complex and unsteady fluid dynamics are not easy to model for simulation tools. One of the most relevant phenomena is the dynamic stall which is inherently to unsteady conditions, this work is focused in its study. 
Dynamic stall is usually used as a passive control for VAWT operating conditions, hence the importance of a good understanding. In this study, \textbf{a coupled model is implemented with the open-source CFD toolbox OpenFOAM for solving the governing equations; an actuator line model is used for solving velocity flow field and a dynamic stall model for force interactions.} Force coefficients obtained from the model are validated with experimental data of pithing airfoil in similar operating conditions as H-rotor type.   
\end{abstract}

\section{Introduction}
Operating conditions of VAWT are characterized for complex unsteady flows which give a considerable challenge for both to describe using measurements and to represent through simulation tools\cite{huyer1996unsteady}. Moreover, VAWT blades are inherently exposed to cyclic variation in the angle of attack which gives cyclic blade forces and can give material fatigue damage. Accurate modeling of the varying forces is therefore very important for the design of VAWT.\\
The angle of attack in a fixed pitch VAWT is increasing with decreased tip speed ration (TSR), and at low TSR (common during stall regulation), the blades will experience dynamic stall, where the force coefficients for the blade not only depend on the angle of attack, but also on the rate of change of the angle of attack. The aim of this study is to investigate the performance of a Leishman-Beddoes type dynamic stall model when implemented within an actuator line model, for pitching motion typical to a VAWT.

\section{Methodology}
For solving the governing equations of phenomena involved, a coupled model have been implemented: the actuator line model is used to calculate the flow velocities and thereby the angle of attack. The dynamic stall model is used to calculate blade force coefficients, which the actuator line model need for the velocity field calculations. In this work it was chosen to validate the model against wind tunnel data for a pitching blade. This will put the focus on the force modeling part in the simulation model\\
\\
\textbf{Actuator Line Model:} is a three-dimensional and unsteady aerodynamic model developed by S{\o}rensen and Shen\cite{sorensen1999computation}, used to study the flow around wind turbines. It is a combination of a solver of Navier-Stokes equations with a so-called actuator line technique, in which blades of the turbine are represented by a radial distribution of body forces along lines. These forces are determined using a dynamic stall model commonly based on empirical data.\\  
\\
\textbf{Dynamic Stall Modeling:} represented by the Leishman-Beddoes model\cite{leishman1986generalised} with the modifications of Sheng et al\cite{sheng2008modified}. It is capable to calculate the unsteady lift, pitching moment and drag, giving a physical description of the aerodynamics. It have been widely validated with experimental data\cite{leishman1989semi}. Separated on three subsystems: an attached flow model for unsteady linear airloads, a separated flow model for non-linear airloads and a dynamic stall model for the airloads induced by the leading edge vortex. In this work, the dynamic stall model described by Dyachuk\cite{dyachuk} has been implemented, using the library turbinesFoam developed by Bachant\cite{bachant2015simulating}.\\
\\
NACA0021 airfoil was tested at the Reynolds number of 1000000 (which is a reasonable value for operating VAWT) during pithing motion similar to the motion of the VAWT blade. Different amplitude degrees were tried out. In a VAWT with fixed blades the angle of attack is in function of the TSR of the turbine. A periodic function is used to represent the variations of the angles of attack which a VAWT blade experiences analog to the pitching motion,
\begin{eqnarray}
\alpha = \arctan \left( \frac{\sin \theta}{\lambda + \cos \theta} \right)
\end{eqnarray}
Where $\theta$ is the azimuthal blade angle and $ \lambda $ represents the TSR, 
\begin{eqnarray}
\lambda =  \frac{\Omega R}{V} 
\end{eqnarray}
The pitching blade experiments were carried out at Glasgow University\cite{angell1988collected}\\

\section{Results and discussion}
The reliability of the model results is acceptable, making a good agreement with experimental results considering the complexity of the flows studied. For a TSR of 3.34 (Figure \ref{fig1}), the maximum magnitude of the angle of attack is at 17.4\degree which is in the stall region. Both CN and CT curves show the delay of the flow reattachment, which is characteristic of the dynamic stall phenomenon. Peaks of simulated values have close agreement with the measured data.\\
For a TSR of 2.60, the maximum amplitude of the angle of attack is 22.6, which is related to a deeper stall region compared to a TSR of 3.34. This is shown because the "loop" of the curve is wider for the lower TSR, and moreover, the reattachment of the flow is further delayed.\\
It have been demonstrated how the coupled actuator line model can be coupled to a modified version of the Leishman-Beddoes model, and the results could be validated for similar operating conditions of a VAWT.

\begin{figure}[h]
\begin{minipage}{18pc}
\includegraphics[width=18pc]{Fig1.png}
\end{minipage}\hspace{2pc}%
\begin{minipage}{18pc}
\includegraphics[width=18pc]{Fig2.png}
\end{minipage} 
\caption{\label{fig1}Normal (left) and tangential (right) force responses during pitching motions of NACA0021 airfoil with a maximum amplitude of 17.4 \degree (analog to a $\lambda$ = 3.34).}
\end{figure}

\begin{figure}[h]
\begin{minipage}{18pc}
\includegraphics[width=18pc]{Fig3.png}
\end{minipage}\hspace{2pc}%
\begin{minipage}{18pc}
\includegraphics[width=18pc]{Fig4.png}
\end{minipage} 
\caption{\label{fig2}Normal (left) and tangential (right) force responses during pitching motions of NACA0021 airfoil with a maximum amplitude of 22.6 \degree (analog to a $\lambda$ = 2.60).}
\end{figure}

\section*{References}
%\begin{thebibliography}{9}
%\bibitem{iopartnum} IOP Publishing is to grateful Mark A Caprio, Center for Theoretical Physics, Yale University, for permission to include the {\tt iopart-num} \BibTeX package (version 2.0, December 21, 2006) with  this documentation. Updates and new releases of {\tt iopart-num} can be found on \verb"www.ctan.org" (CTAN). 
%\end{thebibliography}

\bibliography{mybib}{}
\bibliographystyle{unsrt}

\end{document}

### 2016-02-29 abstract.tex
Use my full name
Wrap abstract text

@@
 \title{Validation of an Actuator Line Model Coupled to a Dynamic Stall Model for Pitching Motion Characteristic to Vertical Axis Turbine} % \\ \jpcs}
-\author{Victor Mendoza$^{1}$, Pete Bachant$^{2}$, Martin Wosnik$^{2}$ and Anders Goude$^{1}$ }
+\author{Victor Mendoza$^{1}$, Peter Bachant$^{2}$, Martin Wosnik$^{2}$ and Anders Goude$^{1}$ }
 \address{$^{1}$ Department of Engineering Sciences, Division of Electricity, Uppsala University, \\Uppsala 751 21, Sweden}
@@
 \begin{abstract}
-Vertical Axis Turbines (VAWT) have been used for extracting of energy either from sea or wind.
-A simpler design, low cost of maintenance and the capability of working independent from the direction of the flow are some of the most important advantages over conventional Horizontal Axis Turbines (HAWTs).
-Complex and unsteady fluid dynamics are not easy to model for simulation tools.
-One of the most relevant phenomena is the dynamic stall which is inherently to unsteady conditions, this work is focused in its study.
-Dynamic stall is usually used as a passive control for VAWT operating conditions, hence the importance of a good understanding.
-In this study, \textbf{a coupled model is implemented with the open-source CFD toolbox OpenFOAM for solving the governing equations; an actuator line model is used for solving velocity flow field and a dynamic stall model for force interactions.} Force coefficients obtained from the model are validated with experimental data of pithing airfoil in similar operating conditions as H-rotor type.
+
+    Vertical Axis Turbines (VAWT) have been used for extracting of energy either
+    from sea or wind. A simpler design, low cost of maintenance and the
+    capability of working independent from the direction of the flow are some of
+    the most important advantages over conventional Horizontal Axis Turbines
+    (HAWTs). Complex and unsteady fluid dynamics are not easy to model for
+    simulation tools. One of the most relevant phenomena is the dynamic stall
+    which is inherently to unsteady conditions, this work is focused in its
+    study. Dynamic stall is usually used as a passive control for VAWT operating
+    conditions, hence the importance of a good understanding. In this study,
+    \textbf{a coupled model is implemented with the open-source CFD toolbox
+    OpenFOAM for solving the governing equations; an actuator line model is used
+    for solving velocity flow field and a dynamic stall model for force
+    interactions.} Force coefficients obtained from the model are validated with
+    experimental data of pithing airfoil in similar operating conditions as
+    H-rotor type.
+
 \end{abstract}

### 2016-02-29 abstract.tex
Wrap text in introduction

@@
 \section{Introduction}
-Operating conditions of VAWT are characterized for complex unsteady flows which give a considerable challenge for both to describe using measurements and to represent through simulation tools\cite{huyer1996unsteady}.
-Moreover, VAWT blades are inherently exposed to cyclic variation in the angle of attack which gives cyclic blade forces and can give material fatigue damage.
-Accurate modeling of the varying forces is therefore very important for the design of VAWT.\\
-The angle of attack in a fixed pitch VAWT is increasing with decreased tip speed ration (TSR), and at low TSR (common during stall regulation), the blades will experience dynamic stall, where the force coefficients for the blade not only depend on the angle of attack, but also on the rate of change of the angle of attack.
-The aim of this study is to investigate the performance of a Leishman-Beddoes type dynamic stall model when implemented within an actuator line model, for pitching motion typical to a VAWT.
+
+Operating conditions of VAWT are characterized for complex unsteady flows which
+give a considerable challenge for both to describe using measurements and to
+represent through simulation tools\cite{huyer1996unsteady}. Moreover, VAWT
+blades are inherently exposed to cyclic variation in the angle of attack which
+gives cyclic blade forces and can give material fatigue damage. Accurate
+modeling of the varying forces is therefore very important for the design of
+VAWT.
+
+The angle of attack in a fixed pitch VAWT is increasing with decreased tip speed
+ration (TSR), and at low TSR (common during stall regulation), the blades will
+experience dynamic stall, where the force coefficients for the blade not only
+depend on the angle of attack, but also on the rate of change of the angle of
+attack. The aim of this study is to investigate the performance of a
+Leishman-Beddoes type dynamic stall model when implemented within an actuator
+line model, for pitching motion typical to a VAWT.
 

### 2016-02-29 paper.tex
Minor formatting edits

--- 2016-02-29 abstract.tex
@@
 \section{Methodology}
-For solving the governing equations of phenomena involved, a coupled model have been implemented: the actuator line model is used to calculate the flow velocities and thereby the angle of attack.
-The dynamic stall model is used to calculate blade force coefficients, which the actuator line model need for the velocity field calculations.
-In this work it was chosen to validate the model against wind tunnel data for a pitching blade.
-This will put the focus on the force modeling part in the simulation model\\
-\\
-\textbf{Actuator Line Model:} is a three-dimensional and unsteady aerodynamic model developed by S{\o}rensen and Shen\cite{sorensen1999computation}, used to study the flow around wind turbines.
-It is a combination of a solver of Navier-Stokes equations with a so-called actuator line technique, in which blades of the turbine are represented by a radial distribution of body forces along lines.
-These forces are determined using a dynamic stall model commonly based on empirical data.\\
-\\
-\textbf{Dynamic Stall Modeling:} represented by the Leishman-Beddoes model\cite{leishman1986generalised} with the modifications of Sheng et al\cite{sheng2008modified}.
-It is capable to calculate the unsteady lift, pitching moment and drag, giving a physical description of the aerodynamics.
-It have been widely validated with experimental data\cite{leishman1989semi}.
-Separated on three subsystems: an attached flow model for unsteady linear airloads, a separated flow model for non-linear airloads and a dynamic stall model for the airloads induced by the leading edge vortex.
-In this work, the dynamic stall model described by Dyachuk\cite{dyachuk} has been implemented, using the library turbinesFoam developed by Bachant\cite{bachant2015simulating}.\\
-\\
-NACA0021 airfoil was tested at the Reynolds number of 1000000 (which is a reasonable value for operating VAWT) during pithing motion similar to the motion of the VAWT blade.
-Different amplitude degrees were tried out.
-In a VAWT with fixed blades the angle of attack is in function of the TSR of the turbine.
-A periodic function is used to represent the variations of the angles of attack which a VAWT blade experiences analog to the pitching motion,
+
+For solving the governing equations of phenomena involved, a coupled model have
+been implemented: the actuator line model is used to calculate the flow
+velocities and thereby the angle of attack. The dynamic stall model is used to
+calculate blade force coefficients, which the actuator line model need for the
+velocity field calculations. In this work it was chosen to validate the model
+against wind tunnel data for a pitching blade. This will put the focus on the
+force modeling part in the simulation model.
+
+\textbf{Actuator Line Model:} is a three-dimensional and unsteady aerodynamic
+model developed by S{\o}rensen and Shen\cite{sorensen1999computation}, used to
+study the flow around wind turbines. It is a combination of a solver of
+Navier-Stokes equations with a so-called actuator line technique, in which
+blades of the turbine are represented by a radial distribution of body forces
+along lines. These forces are determined using a dynamic stall model commonly
+based on empirical data.
+
+\textbf{Dynamic Stall Modeling:} represented by the Leishman-Beddoes
+model\cite{leishman1986generalised} with the modifications of Sheng et
+al\cite{sheng2008modified}. It is capable to calculate the unsteady lift,
+pitching moment and drag, giving a physical description of the aerodynamics. It
+have been widely validated with experimental data\cite{leishman1989semi}.
+Separated on three subsystems: an attached flow model for unsteady linear
+airloads, a separated flow model for non-linear airloads and a dynamic stall
+model for the airloads induced by the leading edge vortex. In this work, the
+dynamic stall model described by Dyachuk\cite{dyachuk} has been implemented,
+using the library turbinesFoam developed by Bachant\cite{bachant2015simulating}.
+
+NACA0021 airfoil was tested at the Reynolds number of 1000000 (which is a
+reasonable value for operating VAWT) during pithing motion similar to the motion
+of the VAWT blade. Different amplitude degrees were tried out. In a VAWT with
+fixed blades the angle of attack is in function of the TSR of the turbine. A
+periodic function is used to represent the variations of the angles of attack
+which a VAWT blade experiences analog to the pitching motion,
 \begin{eqnarray}
@@
 \end{eqnarray}
-The pitching blade experiments were carried out at Glasgow University\cite{angell1988collected}\\
+The pitching blade experiments were carried out at Glasgow
+University\cite{angell1988collected}\\
 
 \section{Results and discussion}
-The reliability of the model results is acceptable, making a good agreement with experimental results considering the complexity of the flows studied.
-For a TSR of 3.34 (Figure \ref{fig1}), the maximum magnitude of the angle of attack is at 17.4\degree which is in the stall region.
-Both CN and CT curves show the delay of the flow reattachment, which is characteristic of the dynamic stall phenomenon.
-Peaks of simulated values have close agreement with the measured data.\\
-For a TSR of 2.60, the maximum amplitude of the angle of attack is 22.6, which is related to a deeper stall region compared to a TSR of 3.34.
-This is shown because the "loop" of the curve is wider for the lower TSR, and moreover, the reattachment of the flow is further delayed.\\
-It have been demonstrated how the coupled actuator line model can be coupled to a modified version of the Leishman-Beddoes model, and the results could be validated for similar operating conditions of a VAWT.
+
+The reliability of the model results is acceptable, making a good agreement with
+experimental results considering the complexity of the flows studied. For a TSR
+of 3.34 (Figure \ref{fig1}), the maximum magnitude of the angle of attack is at
+17.4\degree which is in the stall region. Both CN and CT curves show the delay
+of the flow reattachment, which is characteristic of the dynamic stall
+phenomenon. Peaks of simulated values have close agreement with the measured
+data.
+
+For a TSR of 2.60, the maximum amplitude of the angle of attack is 22.6, which
+is related to a deeper stall region compared to a TSR of 3.34. This is shown
+because the ``loop'' of the curve is wider for the lower TSR, and moreover, the
+reattachment of the flow is further delayed.
+
+It has been demonstrated how the coupled actuator line model can be coupled to
+a modified version of the Leishman-Beddoes model, and the results could be
+validated for similar operating conditions of a VAWT.
 

### 2016-03-01 paper.tex
Tweak title a little
Remove some whitespace
New version - A. Goude suggestions
Anders suggested modifications

@@
 
-\title{Validation of an Actuator Line Model Coupled to a Dynamic Stall Model for Pitching Motion Characteristic to Vertical Axis Turbine} % \\ \jpcs}
+\title{Validation of an Actuator Line Model Coupled to a Dynamic Stall Model for
+Pitching Motions Characteristic to Vertical Axis Turbines}
+
 \author{Victor Mendoza$^{1}$, Peter Bachant$^{2}$, Martin Wosnik$^{2}$ and Anders Goude$^{1}$ }
@@
 
-\begin{abstract}
-
-    Vertical Axis Turbines (VAWT) have been used for extracting of energy either
+\begin{abstract}Vertical Axis Wind Turbines (VAWTs) have been used for extracting of energy either
     from sea or wind. A simpler design, low cost of maintenance and the
     capability of working independent from the direction of the flow are some of
-    the most important advantages over conventional Horizontal Axis Turbines
+    the most important advantages over conventional Horizontal Axis Wind Turbines
     (HAWTs). Complex and unsteady fluid dynamics are not easy to model for
@@
     experimental data of pithing airfoil in similar operating conditions as
-    H-rotor type.
+    H-rotor type. Numerical results show reasonable agreement with experimental data
+	for pitching motion. In general, simulations allow to characterize dynamic stall effects 
+	and their magnitude.
 
@@
 
-Operating conditions of VAWT are characterized for complex unsteady flows which
-give a considerable challenge for both to describe using measurements and to
+Operating conditions of VAWT are characterized as complex unsteady flows which
+give a considerable challenge, both to describe using measurements and to
 represent through simulation tools\cite{huyer1996unsteady}. Moreover, VAWT
-blades are inherently exposed to cyclic variation in the angle of attack which
+blades are inherently exposed to cyclic variation in the angle of attack, which
 gives cyclic blade forces and can give material fatigue damage. Accurate
-modeling of the varying forces is therefore very important for the design of
+modeling of the varying forces is therefore very important for the design of the
 VAWT.
 
-The angle of attack in a fixed pitch VAWT is increasing with decreased tip speed
+The amplitude of the angle of attack oscillation in a fixed pitch VAWT is increasing 
+with decreased tip speed
 ration (TSR), and at low TSR (common during stall regulation), the blades will
@@
 
-For solving the governing equations of phenomena involved, a coupled model have
+For solving the governing equations of the phenomena involved, a coupled model have
 been implemented: the actuator line model is used to calculate the flow
 velocities and thereby the angle of attack. The dynamic stall model is used to
-calculate blade force coefficients, which the actuator line model need for the
-velocity field calculations. In this work it was chosen to validate the model
+calculate blade force coefficients, which the actuator line model needs for the
+velocity field calculations.
+
+In this work, it was chosen to validate the model
 against wind tunnel data for a pitching blade. This will put the focus on the
@@
 model\cite{leishman1986generalised} with the modifications of Sheng et
-al\cite{sheng2008modified}. It is capable to calculate the unsteady lift,
+al\cite{sheng2008modified} and Dyachuk\cite{dyachuk},
+using the library turbinesFoam developed by Bachant\cite{bachant2015simulating}.
+It is capable to calculate the unsteady lift,
 pitching moment and drag, giving a physical description of the aerodynamics. It
-have been widely validated with experimental data\cite{leishman1989semi}.
-Separated on three subsystems: an attached flow model for unsteady linear
+have been validated with experimental data in\cite{leishman1989semi}.
+The model is separated into three subsystems: an attached flow model for unsteady linear
 airloads, a separated flow model for non-linear airloads and a dynamic stall
-model for the airloads induced by the leading edge vortex. In this work, the
-dynamic stall model described by Dyachuk\cite{dyachuk} has been implemented,
-using the library turbinesFoam developed by Bachant\cite{bachant2015simulating}.
+model for the airloads induced by the leading edge vortex.
 
-NACA0021 airfoil was tested at the Reynolds number of 1000000 (which is a
+A NACA0021 airfoil was tested at the Reynolds number of 1.000.000 (which is a
 reasonable value for operating VAWT) during pithing motion similar to the motion
-of the VAWT blade. Different amplitude degrees were tried out. In a VAWT with
-fixed blades the angle of attack is in function of the TSR of the turbine. A
+of the VAWT blade. Different amplitude degrees were investigated. In a VAWT with
+fixed blades, the angle of attack is in function of the TSR of the turbine. A
 periodic function is used to represent the variations of the angles of attack
@@
 
-\section{Results and discussion}
-
-The reliability of the model results is acceptable, making a good agreement with
-experimental results considering the complexity of the flows studied. For a TSR
-of 3.34 (Figure \ref{fig1}), the maximum magnitude of the angle of attack is at
-17.4\degree which is in the stall region. Both CN and CT curves show the delay
-of the flow reattachment, which is characteristic of the dynamic stall
-phenomenon. Peaks of simulated values have close agreement with the measured
-data.
-
-For a TSR of 2.60, the maximum amplitude of the angle of attack is 22.6, which
-is related to a deeper stall region compared to a TSR of 3.34. This is shown
-because the ``loop'' of the curve is wider for the lower TSR, and moreover, the
-reattachment of the flow is further delayed.
-
-It has been demonstrated how the coupled actuator line model can be coupled to
-a modified version of the Leishman-Beddoes model, and the results could be
-validated for similar operating conditions of a VAWT.
-
 \begin{figure}[h]
@@
 \end{minipage}
-\caption{\label{fig1}Normal (left) and tangential (right) force responses during pitching motions of NACA0021 airfoil with a maximum amplitude of 17.4 \degree (analog to a $\lambda$ = 3.34).}
+\caption{\label{fig1}Normal (left) and tangential (right) force responses during pitching motions of NACA0021 airfoil with a maximum amplitude of 17.4\degree\ (analog to a $\lambda$ = 3.34).}
 \end{figure}
@@
 \end{minipage}
-\caption{\label{fig2}Normal (left) and tangential (right) force responses during pitching motions of NACA0021 airfoil with a maximum amplitude of 22.6 \degree (analog to a $\lambda$ = 2.60).}
+\caption{\label{fig2}Normal (left) and tangential (right) force responses during pitching motions of NACA0021 airfoil with a maximum amplitude of 22.6\degree\ (analog to a $\lambda$ = 2.60).}
 \end{figure}
+
+\newpage
+
+\section{Results and discussion}
+
+For a TSR of 3.34 (Figure \ref{fig1}), the maximum magnitude of the angle of attack is at
+17.4\degree\ which is in the stall region. Both CN and CT curves show the delay
+of the flow reattachment, which is characteristic of the dynamic stall
+phenomenon. Peaks of simulated values have close agreement with the measured
+data.
+
+For a TSR of 2.60 (Figure \ref{fig2}), the maximum amplitude of the angle of attack is 22.6\degree, which
+is related to a deeper stall region compared to a TSR of 3.34. This is shown
+because the ``loop'' of the curve is wider for the lower TSR, and moreover, the
+reattachment of the flow is further delayed.
+
+The coupled model was set-up for the operating conditions of a VAWT. The reliability of the results is acceptable, making a good agreement with
+experimental data from the wind tunnel test, considering the complexity of the flows studied.
+
+%It has been demonstrated how the coupled actuator line model can be coupled to
+%a modified version of the Leishman-Beddoes model, and the results could be
+%validated for similar operating conditions of a VAWT.
 

### 2016-03-01 paper.tex
Wrapping and whitespace

@@
 
-\begin{abstract}Vertical Axis Wind Turbines (VAWTs) have been used for extracting of energy either
-    from sea or wind. A simpler design, low cost of maintenance and the
+\begin{abstract}
+
+    Vertical Axis Wind Turbines (VAWTs) have been used for extracting of energy
+    either from sea or wind. A simpler design, low cost of maintenance and the
     capability of working independent from the direction of the flow are some of
-    the most important advantages over conventional Horizontal Axis Wind Turbines
-    (HAWTs). Complex and unsteady fluid dynamics are not easy to model for
-    simulation tools. One of the most relevant phenomena is the dynamic stall
-    which is inherently to unsteady conditions, this work is focused in its
-    study. Dynamic stall is usually used as a passive control for VAWT operating
-    conditions, hence the importance of a good understanding. In this study,
-    \textbf{a coupled model is implemented with the open-source CFD toolbox
-    OpenFOAM for solving the governing equations; an actuator line model is used
-    for solving velocity flow field and a dynamic stall model for force
+    the most important advantages over conventional Horizontal Axis Wind
+    Turbines (HAWTs). Complex and unsteady fluid dynamics are not easy to model
+    for simulation tools. One of the most relevant phenomena is the dynamic
+    stall which is inherently to unsteady conditions, this work is focused in
+    its study. Dynamic stall is usually used as a passive control for VAWT
+    operating conditions, hence the importance of a good understanding. In this
+    study, \textbf{a coupled model is implemented with the open-source CFD
+    toolbox OpenFOAM for solving the governing equations; an actuator line model
+    is used for solving velocity flow field and a dynamic stall model for force
     interactions.} Force coefficients obtained from the model are validated with
     experimental data of pithing airfoil in similar operating conditions as
-    H-rotor type. Numerical results show reasonable agreement with experimental data
-	for pitching motion. In general, simulations allow to characterize dynamic stall effects 
-	and their magnitude.
+    H-rotor type. Numerical results show reasonable agreement with experimental
+    data for pitching motion. In general, simulations allow to characterize
+    dynamic stall effects  and their magnitude.
 

### 2016-03-02 paper.tex
New corrections
New corrections Goude - Mendoza

@@
     H-rotor type. Numerical results show reasonable agreement with experimental
-    data for pitching motion. In general, simulations allow to characterize
-    dynamic stall effects  and their magnitude.
+    data for pitching motion.
 
@@
 
-Operating conditions of VAWT are characterized as complex unsteady flows which
+Operating conditions of VAWTs are characterized as complex unsteady flows which
 give a considerable challenge, both to describe using measurements and to
@@
 along lines. These forces are determined using a dynamic stall model commonly
-based on empirical data.
+based on empirical data. This work uses the library turbinesFoam developed by
+Bachant\cite{bachant2015simulating}.
 
-\textbf{Dynamic Stall Modeling:} represented by the Leishman-Beddoes
+\textbf{Dynamic Stall Modeling:} is represented by the Leishman-Beddoes
 model\cite{leishman1986generalised} with the modifications of Sheng et
-al\cite{sheng2008modified} and Dyachuk\cite{dyachuk},
-using the library turbinesFoam developed by Bachant\cite{bachant2015simulating}.
+al\cite{sheng2008modified} and Dyachuk\cite{dyachuk}.
 It is capable to calculate the unsteady lift,
@@
 reasonable value for operating VAWT) during pithing motion similar to the motion
-of the VAWT blade. Different amplitude degrees were investigated. In a VAWT with
+of the VAWT blade. Different pitching amplitudes were investigated. In a VAWT with
 fixed blades, the angle of attack is in function of the TSR of the turbine. A
@@
 \begin{eqnarray}
-    \alpha = \arctan \left( \frac{\sin \theta}{\lambda + \cos \theta} \right)
+    \alpha = \arctan \left( \frac{\sin \theta}{\lambda + \cos \theta} \right),
 \end{eqnarray}
-Where $\theta$ is the azimuthal blade angle and $ \lambda $ represents the TSR,
+where $\theta$ is the azimuthal blade angle and $ \lambda $ represents the TSR,
 \begin{eqnarray}
-    \lambda =  \frac{\Omega R}{V}
+    \lambda =  \frac{\Omega R}{V}.
 \end{eqnarray}
 The pitching blade experiments were carried out at Glasgow
-University\cite{angell1988collected}\\
+University\cite{angell1988collected}.\\
 
@@
 \end{minipage}
-\caption{\label{fig1}Normal (left) and tangential (right) force responses during pitching motions of NACA0021 airfoil with a maximum amplitude of 17.4\degree\ (analog to a $\lambda$ = 3.34).}
+\caption{\label{fig1}Normal (left) and tangential (right) force coefficients during pitching motions of NACA0021 airfoil with a maximum amplitude of 17.4\degree\ (analog to $\lambda = 3.34$).}
 \end{figure}
@@
 \end{minipage}
-\caption{\label{fig2}Normal (left) and tangential (right) force responses during pitching motions of NACA0021 airfoil with a maximum amplitude of 22.6\degree\ (analog to a $\lambda$ = 2.60).}
+\caption{\label{fig2}Normal (left) and tangential (right) force coefficients during pitching motions of NACA0021 airfoil with a maximum amplitude of 22.6\degree\ (analog to $\lambda = 2.60$).}
 \end{figure}
@@
 For a TSR of 3.34 (Figure \ref{fig1}), the maximum magnitude of the angle of attack is at
-17.4\degree\ which is in the stall region. Both CN and CT curves show the delay
+17.4\degree\ which is in the stall region. Both \textit{CN} and \textit{CT} curves show the delay
 of the flow reattachment, which is characteristic of the dynamic stall
@@
 
-The coupled model was set-up for the operating conditions of a VAWT. The reliability of the results is acceptable, making a good agreement with
-experimental data from the wind tunnel test, considering the complexity of the flows studied.
-
-%It has been demonstrated how the coupled actuator line model can be coupled to
-%a modified version of the Leishman-Beddoes model, and the results could be
-%validated for similar operating conditions of a VAWT.
+In general, the model show good agreement with experimental data, considering the complexity
+of the flows studied. This makes the model suitable for application in wind turbine
+simulations.
 

### 2016-03-06 paper.tex
new script with new figures
Clean up abstract wording a bit

@@
 
-    Vertical Axis Wind Turbines (VAWTs) have been used for extracting of energy
-    either from sea or wind. A simpler design, low cost of maintenance and the
-    capability of working independent from the direction of the flow are some of
-    the most important advantages over conventional Horizontal Axis Wind
-    Turbines (HAWTs). Complex and unsteady fluid dynamics are not easy to model
-    for simulation tools. One of the most relevant phenomena is the dynamic
-    stall which is inherently to unsteady conditions, this work is focused in
-    its study. Dynamic stall is usually used as a passive control for VAWT
-    operating conditions, hence the importance of a good understanding. In this
-    study, \textbf{a coupled model is implemented with the open-source CFD
-    toolbox OpenFOAM for solving the governing equations; an actuator line model
-    is used for solving velocity flow field and a dynamic stall model for force
-    interactions.} Force coefficients obtained from the model are validated with
+    Vertical axis turbines (VATs) can be used to extract renewable energy from
+    wind or water flows. A simpler design, low cost of maintenance, and the
+    ability to accept flow from all directions perpendiculat to the rotor axis
+    are some of the most important advantages over conventional horizontal axis
+    turbines (HATs). However, VATs encounter complex and unsteady fluid
+    dynamics, which present significant modeling challenges. One of the most
+    relevant phenomena is dynamic stall, which is caused by the unsteady
+    variation of angle of attack throughout the blade rotation, and is the focus
+    of the present study. Dynamic stall is usually used as a passive control for
+    VAT operating conditions, hence the importance of predicting its effects. In
+    this study, \textbf{a coupled model is implemented with the open-source CFD
+    toolbox OpenFOAM for solving the Navier--Stokes equations, where an actuator
+    line model and dynamic stall model are used compute the blade loading and
+    body force.} Force coefficients obtained from the model are validated with
     experimental data of pithing airfoil in similar operating conditions as
@@
 \begin{minipage}{18pc}
-\includegraphics[width=18pc]{Fig1.png}
+%\includegraphics[width=18pc]{Fig1.png}
+\resizebox{\columnwidth}{!}{\input{CN1}}
 \end{minipage}\hspace{2pc}%
 \begin{minipage}{18pc}
-\includegraphics[width=18pc]{Fig2.png}
+%\includegraphics[width=18pc]{Fig2.png}
+\resizebox{\columnwidth}{!}{\input{CT1}}
 \end{minipage}
@@
 \begin{minipage}{18pc}
-\includegraphics[width=18pc]{Fig3.png}
+%\includegraphics[width=18pc]{Fig3.png}
+\resizebox{\columnwidth}{!}{\input{CN2}}
 \end{minipage}\hspace{2pc}%
 \begin{minipage}{18pc}
-\includegraphics[width=18pc]{Fig4.png}
+%\includegraphics[width=18pc]{Fig4.png}
+\resizebox{\columnwidth}{!}{\input{CT2}}
 \end{minipage}

### 2016-03-06 paper.tex
Clarify that an H-rotor is a VAT

@@
     body force.} Force coefficients obtained from the model are validated with
-    experimental data of pithing airfoil in similar operating conditions as
-    H-rotor type. Numerical results show reasonable agreement with experimental
-    data for pitching motion.
+    experimental data of pithing airfoil in similar operating conditions as an
+    H-rotor type VAT. Numerical results show reasonable agreement with
+    experimental data for pitching motion.
 

### 2016-03-06 paper.tex
corrected spelling

@@
     wind or water flows. A simpler design, low cost of maintenance, and the
-    ability to accept flow from all directions perpendiculat to the rotor axis
+    ability to accept flow from all directions perpendicular to the rotor axis
     are some of the most important advantages over conventional horizontal axis

### 2016-06-03 paper.tex
Add newest files from @victormendoza83

@@
 \usepackage{gensymb}
+\usepackage{epstopdf}
+\usepackage{amsmath}
+\usepackage{amssymb} 	%for symbols
+
+\graphicspath{{./figures/}}
 
 \begin{document}
-
 \title{Validation of an Actuator Line Model Coupled to a Dynamic Stall Model for
@@
 \begin{abstract}
-
     Vertical axis turbines (VATs) can be used to extract renewable energy from
@@
     VAT operating conditions, hence the importance of predicting its effects. In
-    this study, \textbf{a coupled model is implemented with the open-source CFD
-    toolbox OpenFOAM for solving the Navier--Stokes equations, where an actuator
-    line model and dynamic stall model are used compute the blade loading and
-    body force.} Force coefficients obtained from the model are validated with
-    experimental data of pithing airfoil in similar operating conditions as an
+    this study, a coupled model is implemented with the open-source CFD
+    toolbox OpenFOAM for solving the Navier-Stokes equations, where an actuator
+    line model and dynamic stall model are used to compute the blade loading and
+    body force. Force coefficients obtained from the model are validated with
+    experimental data of pitching airfoil in similar operating conditions as an
     H-rotor type VAT. Numerical results show reasonable agreement with
     experimental data for pitching motion.
-
 \end{abstract}
@@
 \section{Introduction}
-
-Operating conditions of VAWTs are characterized as complex unsteady flows which
+Operating conditions of Vertical Axis Wind Turbines VAWTs are characterized as complex unsteady flows which
 give a considerable challenge, both to describe using measurements and to
@@
 \section{Methodology}
-
 For solving the governing equations of the phenomena involved, a coupled model have
@@
 
-\textbf{Actuator Line Model:} is a three-dimensional and unsteady aerodynamic
+\subsection{Actuator Line Model (ALM)} Is a three-dimensional and unsteady aerodynamic
 model developed by S{\o}rensen and Shen\cite{sorensen1999computation}, used to
@@
 based on empirical data. This work uses the library turbinesFoam developed by
-Bachant\cite{bachant2015simulating}.
+Bachant\cite{bachant2015simulating}. This model is based on incompressible Navier-Stokes equation
+\begin{align}
+ & \nonumber \\ & \dfrac{\partial \rho \vec{V}}{\partial t} + \vec{V} \cdot \nabla \vec{V} = - \dfrac{1}{ \rho} \nabla p + \nu \nabla ^2 \vec{V} + \vec{f}	  \label{eqmom}  
+\end{align} %\begin{align}
+with $\vec{f}$ representing the body forces, which are the loads on the rotating blades for this case.
 
-\textbf{Dynamic Stall Modeling:} is represented by the Leishman-Beddoes
+Lift and drag coefficients of a 2D profile are already known, and these are tabulated and combined with the blade element approach for determining 
+body forces acting on blades.
+Relative flow velocity $\vec{V}_{rel} $ and angle of attack $\alpha$ are calculated for each blade using the geometric relation between the velocity of the blade $- \omega r$, with $\omega$ representing the angular (rotational) velocity of the rotor and $r$ is the radius of the blade element, and the velocity of the local incoming wind flow $\vec{V}_{in}$.
+Figure \ref{figvectors} depicts a cross-sectional airfoil element at radius $r$ in the plane perpendicular to the turbine axis.
+
+\begin{figure}[h]
+%\begin{minipage}{36pc}
+\begin{center}
+\includegraphics[width=0.35\columnwidth]{vector.eps}
+\end{center}
+%\resizebox{\columnwidth}{!}{\input{plot}}
+%\end{minipage}%
+\caption{\label{figvectors} Illustration of velocity vectors and forces acting at the cross-section airfoil element}
+\end{figure}
+
+Once the angle of attack, relative velocity are obtained, the blade element lift and drag forces per spanwise length unit are calculated as
+\begin{align}
+& \vec{f}_L = \frac{1}{2} \ \rho \ c \ C_L \left| \vec{V}_{rel}^2 \right|  \label{lift}  
+\end{align} %\begin{align}
+\begin{align}
+& \vec{f}_D = \frac{1}{2} \ \rho \ c \ C_D \left| \vec{V}_{rel}^2 \right|  \label{drag}  
+\end{align} %\begin{align}
+where $C_L$ and $C_D$ are the lift and drag coefficients, respectively.
+Both are in function of the Reynolds number and the angle of attack.
+It should be emphasized that the direction of the lift and drag are respectively perpendicular and parallel to the relative velocity $\vec{V}_{rel}$.
+$c$ represents the chord length.
+
+Same procedure is used to calculate forces from the turbine shaft and blade support struts.
+After all these forces on the actuator lines are calculated, then are added as a body force source, per unit of density (incompressible assumption), in the momentum Navier-Stokes equations (equation \ref{eqmom}).
+
+\subsection{Dynamic Stall Modeling (DSM)} Is represented by the Leishman-Beddoes
 model\cite{leishman1986generalised} with the modifications of Sheng et
@@
 
+The unsteady attached flow solution considers the circulatory and impulsive loads, which, at the same time depend on the unsteady bound vortex.
+The circulatory normal force coefficient could be expressed as
+\begin{align}
+& C_{N_n}^C = C_{N_\alpha} \alpha_{E_n}	\label{CNcirc}
+\end{align}
+where $C_{N \alpha}$ indicates the slope of the static normal force coefficient for a specific Reynolds number, and $\alpha_{E_n}$ is an expression for an equivalent angle of attack
+\begin{align}
+& \alpha_{E_n} = \alpha_n - X_n -Y_n - Z_n	\label{alphaeq}
+\end{align}
+where $\alpha$ is the geometrical angle of attackm $X$, $Y$ and $Z$ are the deficiency functions, which are empirically derived, based on the flow velocity and the pithing rate, the details are found in \cite{dyachuk2013dynamic}.
+Indices $n$ and $n-1$ in equations \ref{CNcirc} and \ref{alphaeq} represent the current and previous time-step.
+A delayed angle of attack is considered due to the lag in pressure response, and it is calculated as
+\begin{align}
+& {\alpha _n}' = {\alpha _n} - D_{\alpha _n}    	\label{alphadel}
+\end{align}
+with $D_\alpha$ as the deficiency function
+\begin{align}
+& D_{\alpha_n} = D_{\alpha_{n-1}} \exp \left( - \frac{\Delta s}{T_\alpha} \right) + (\alpha_n - \alpha_{n-1})\exp \left( - \frac{\Delta s}{2T_\alpha} \right) \label{defalpha}
+\end{align}
+with the empirically derived time constant $T_\alpha$, which has a value of $T_\alpha$ = 6.3 for the NACA0021 airfoil, and $\Delta s$ corresponds to a non-dimensional time-step
+\begin{align}
+& \Delta s| = \frac{2 | \vec{V}_{rel} | \Delta t}{c}		.	\label{deltas}
+\end{align}
+
+Because of the reversal flow in the boundary layer, a leading edge vortex is created at the surface of the airfoil.
+The critical angle of attack $\alpha_{cr_n}$ defines the condition at which dynamic stall begins
+\begin{align}
+\alpha_{cr_n} =
+\begin{cases}
+    \alpha_{ds0}       & \quad |r_n| \geqslant r_0 \\
+    \alpha_{ss} + (\alpha_{ds0} - \alpha_{ss}) \frac{| r_n |}{r_0}       & \quad |r_n| < r_0 \\
+\end{cases}		\label{criticalpha}
+\end{align}
+with the reduced pitching rate $r_n$ defined as
+\begin{align}
+r_n = \frac{\dot{\alpha}_n c}{ 2 | \vec{V}_{rel} | }
+\end{align}
+here, $\dot{\alpha}$ is the pitch rate, $r_0$ is the reduced pitching rate, which limits the quasi-steady stall with the dynamic stall.
+Its value is $r_0$ = 0.01 for the NACA0021 airfoil.
+The static stall onset angle $\alpha_{ss}$ and the critical stall onset angle $\alpha_{ds0}$ are 14.33 and 17.91, respectively.
+The following is the the dynamic stall condition used
+\begin{align}
+| \alpha ' | > \alpha_{cr} \rightarrow \mathrm{stall} \label{stall}
+\end{align}
+for the separated flow, the effects are divided in two groups: trailing edge separation and leading vortex convection.
+The first one is related to the time delay in the movement of the boundary separation point, and is calculated using Kirchhoff's approximation
+\begin{align}
+f'_n =
+\begin{cases}
+    1- 0.4 \exp \left( \frac{|\alpha'_n| - \alpha_1 }{S_1}   \right)    & \quad |\alpha'_n| <  \alpha_1 \\
+    0.02- 0.58 \exp \left( \frac{\alpha_1 - |\alpha'_n| }{S_2} \right)  & \quad |\alpha'_n| \geqslant  \alpha_1 \\
+\end{cases}		\label{criticalpha2}
+\end{align}
+where $f'$ is the delayed separation point and $\alpha_1$, $S_1$ and $S_2$ are constant based on the airfoil profile and the local Reynolds number, these can be found in \cite{dyachuk2013dynamic}.
+The boundary layer around the blade is in function on the time, and this effect is superimposed on the pressure response delay.
+The additional delay is represented by $f''$, which is the dynamic separation point
+\begin{align}
+f''_n = f'_n - D _{f_n}	   \label{dyndelay}
+\end{align}
+and $D_{f_n}$ is the deficiency function defined as
+\begin{align}
+& D_{f_n} = D_{f_{n-1}} \exp \left( - \frac{\Delta s}{T_f} \right) + (f_n - f_{n-1})\exp \left( - \frac{\Delta s}{2T_f} \right) \label{deff}
+\end{align}
+here, $T_f$ is empirically derived time constant and its value is $T_f$ = 3. Then, the normal force coefficient for unsteady conditions before the dynamic stall onset is calculated as
+\begin{align}
+& C_{N_n}^f = C_{N_\alpha} \alpha_{E_n} \left( \frac{1 + \sqrt{f''}}{2} \right)^2	\label{CNf}
+\end{align}
+after the stall condition is met, the leading edge vortex convects over the surface of the airfoil towards the trailing edge. A significant increase in the normal forces occurs during this process
+\begin{align}
+C_N^v = B_1 (f'' - f) V_x \label{CNv}
+\end{align}
+where $C_N^v$ represents the normal force coefficient during the vortex convection, which is in function of the pitch rate.
+$V_x$ and $B_1$ are parameters based on the local Reynolds number and the airfoil profile, they are found in \cite{dyachuk2013dynamic}.
+Once the vortex passes the trailing edge, the normal force decreases quickly.
+Total normal force obtained is expressed as
+\begin{align}
+C_N = C_{N_n}^f + C_N^v \label{CNtotal}
+\end{align}
+Figure \ref{figpitching} shows an example of the normal coefficient forces on a blade during pitching motion, using the above model described.
+
+\begin{figure}[h]
+%\begin{minipage}{36pc}
+\begin{center}
+\includegraphics[width=0.5\columnwidth]{CNpitching2.eps}
+\end{center}
+%\resizebox{\columnwidth}{!}{\input{plot}}
+%\end{minipage}%
+\caption{\label{figpitching} Illustration of dynamic stall: NACA0021 profile with $\alpha=6 + 16 \sin (\omega	t), \ \omega=12,47$ [rad/s], $V_\infty=34,3$ [m/s] and $c$ = 0,55[m]}
+\end{figure}
+
+Tangential force coefficient is calculated trough Kichhoff's flow relation using the dynamic separation point.
+\begin{align}
+C_T = \eta C_{N_\alpha} \alpha_E ^2 ( \sqrt{f''}-E_0)     \label{CT}
+\end{align}
+with the empirical constants $\eta = 0.975$ and $E_0 = 0.15$ corresponding to NACA0021 airfoil profile.
+
+\section{Simulation Parameters: Validation case}
 A NACA0021 airfoil was tested at the Reynolds number of 1.000.000 (which is a
-reasonable value for operating VAWT) during pithing motion similar to the motion
-of the VAWT blade. Different pitching amplitudes were investigated. In a VAWT with
+reasonable value for operating VAWT) during pitching motion similar to the motion
+of the VAWT blade.
+Different pitching amplitudes were investigated: 13.8, 17.4 and 22.6$\degree$, analog to a TSR of $\lambda=$ 4.19, 3.34 and 2.6 respectively (see equation \ref{tsrval}), to cover a wide range of operational conditions of a VAWT.
+In a VAWT with
 fixed blades, the angle of attack is in function of the TSR of the turbine. A
@@
 \begin{eqnarray}
-    \alpha = \arctan \left( \frac{\sin \theta}{\lambda + \cos \theta} \right),
+    \alpha = \arctan \left( \frac{\sin \theta}{\lambda + \cos \theta} \right) \label{alphaval}
 \end{eqnarray}
@@
 \begin{eqnarray}
-    \lambda =  \frac{\Omega R}{V}.
+    \lambda =  \frac{\Omega R}{ |\vec{V}_\infty | } \label{tsrval}
 \end{eqnarray}
+In the equation \ref{alphaval}, $\alpha$ represents the geometric angle of attack, and it is different to the effective angle of attack $\alpha_E$, which is calculated by the model in order to get the force coefficients.
+A NACA0021 profile with 0.55 [m] of chord length is driven to the dynamic stall region trough the pitching motion accordint to equation \ref{alphaval}.
 The pitching blade experiments were carried out at Glasgow
-University\cite{angell1988collected}.\\
+University\cite{angell1988collected}.
+
+\section{Results and discussion}
+In this section, obtained results for $C_N$ and $C_T$ from simulations are compared with the experimental data.
+Varying parameters have been tested, mainly in the deep stall condition for an amplitude of 22.6 $\degree $ (equivalent to $\lambda$ = 2.6).
+All the studied cases were simulated using a LES Smagorinsky turbulence model.
+Figure \ref{revolutions} shows that the model needs two revolutions to reach the convergence, a third revolution will not produce any change.
 
@@
 \begin{minipage}{18pc}
-%\includegraphics[width=18pc]{Fig1.png}
-\resizebox{\columnwidth}{!}{\input{CN1}}
+\includegraphics[width=\columnwidth]{CN226revolutions.eps}
+%\resizebox{\columnwidth}{!}{\input{plot}}
 \end{minipage}\hspace{2pc}%
 \begin{minipage}{18pc}
-%\includegraphics[width=18pc]{Fig2.png}
-\resizebox{\columnwidth}{!}{\input{CT1}}
+\includegraphics[width=\columnwidth]{CT226revolutions.eps}
 \end{minipage} 
-\caption{\label{fig1}Normal (left) and tangential (right) force coefficients during pitching motions of NACA0021 airfoil with a maximum amplitude of 17.4\degree\ (analog to $\lambda = 3.34$).}
+\caption{\label{revolutions}Normal (left) and tangential (right) force coefficients varying the number of revolutions for a pitching motion of NACA0021 airfoil with a maximum amplitude of 22.6$\degree$.}
+\end{figure}
+
+\textbf{Spatial sensitivity:} A test of the response of the model 
+to the variation in the size of the mesh is carried out.
+A mesh of approximatively 1e5 of cells with a local refinement around the blade is used as a reference mesh (Figure \ref{figrefmesh}), this topology is kept constant and globally refinement are applied for this study; it means the mesh is proportionally scaled in every coordinates.
+
+\begin{figure}[h]
+\begin{minipage}{36pc}
+\includegraphics[width=0.5\columnwidth]{vorticityZ2.eps}
+\includegraphics[width=0.5\columnwidth]{vorticityZ.png}
+%\resizebox{\columnwidth}{!}{\input{plot}}
+\end{minipage}%
+\caption{\label{figrefmesh} Illustration of a reference mesh section and vorticity isosurfaces generated by the pitching motion around the blade.}
+\end{figure}
+
+From Figure \ref{figmesh}, it could be noticed that there is a  variation in the results either for a decreasing or increasing in the mesh resolution.
+Between the refinement of 150 and 200\% not considerable changes are obtained in the results, so a major refinement is not justified due this will not give any possible improvement in the accuracy of the results.
+
+\begin{figure}[h]
+\begin{minipage}{18pc}
+\includegraphics[width=\columnwidth]{CNmesh.eps}
+%\resizebox{\columnwidth}{!}{\input{plot}}
+\end{minipage}\hspace{2pc}%
+\begin{minipage}{18pc}
+\includegraphics[width=\columnwidth]{CTmesh.eps}
+\end{minipage} 
+\caption{\label{figmesh}Normal (left) and tangential (right) force coefficients varying the size of the reference mesh for a pitching motion with a maximum amplitude of 22.6\degree\ }
+\end{figure}
+
+\textbf{Temporal sensitivity:} Different time-steps were chosen for a varying temporal discretization test, which $\Delta t$ values are 0.001, 0.003, 0.0055 and 0.0075 [s], and these correspond to around 1200, 400, 220 and 160 time-steps per revolution respectively.
+In this study a coarse mesh with a refinement of 25\%, compared to the reference one, was used.
+Results are shown in Figure \ref{figdtcoarse}.
+
+The model is more sensitive to the spatial discretization than the temporal one; same characteristic was obtained by Bachant et al.
+in \cite{Bachant2016-VAT-ALM}.
+Accuracy and stability were not affected even for big values for $\Delta t$, as long they are within the limit given by the CFL condition; a Courant number of less than unity is required.
+On the other hand, it should be considered that small time-step of discretization could cause numerical instabilities due to the fluctuation of the flow fields resolving the transient term $\frac{\partial}{\partial t}$.
+
+\begin{figure}[h]
+\begin{minipage}{18pc}
+\includegraphics[width=\columnwidth]{CN226dtLEScoarse.eps}
+%\resizebox{\columnwidth}{!}{\input{plot}}
+\end{minipage}\hspace{2pc}%
+\begin{minipage}{18pc}
+\includegraphics[width=\columnwidth]{CT226dtLEScoarse.eps}
+\end{minipage} 
+\caption{\label{figdtcoarse}Normal (left) and tangential (right) force coefficients varying time-step discretization for a pithing motion with a maximum amplitude of 22.6\degree\ }
+\end{figure}
+
+\subsection{TSR Variation}
+
+For a TSR of 4.19 (Figure \ref{fig138}), peaks values for both $C_N$ and $C_T$ are similar to experiments.
+In this condition, the flow is attached and the blades are not in the stall region.
+Therefore, model have good agreement with experimental data using unsteady attached angle of attack calculation.
+
+For a TSR of 3.34 (Figure \ref{fig174}), the maximum magnitude of the angle of attack is at
+17.4\degree\ which is in the stall region. Both \textit{CN} and \textit{CT} curves show the delay
+of the flow reattachment, which is characteristic of the dynamic stall
+phenomenon. Peaks of simulated values have close agreement with the measured
+data. The stall onset angle flow reattachment are well predicted.
+
+For a TSR of 2.60 (Figure \ref{fig226}), the maximum amplitude of the angle of attack is 22.6\degree, which
+is related to a deeper stall region compared to a TSR of 3.34. This is shown
+because the ``loop'' of the curve is wider for the lower TSR, and moreover, the
+reattachment of the flow is further delayed. Simulated peak for $C_T$ is overestimated.
+
+\begin{figure}[h]
+\begin{minipage}{18pc}
+\includegraphics[width=\columnwidth]{CN138.eps}
+%\resizebox{\columnwidth}{!}{\input{plot}}
+\end{minipage}\hspace{2pc}%
+\begin{minipage}{18pc}
+\includegraphics[width=\columnwidth]{CT138.eps}
+\end{minipage} 
+\caption{\label{fig138}Normal (left) and tangential (right) force coefficients during pitching motions of NACA0021 airfoil with a maximum amplitude of 13.8\degree\ (analog to $\lambda = 4.19$).}
 \end{figure}
@@
 \begin{minipage}{18pc}
-%\includegraphics[width=18pc]{Fig3.png}
-\resizebox{\columnwidth}{!}{\input{CN2}}
+\includegraphics[width=\columnwidth]{CN174.eps}
+%\resizebox{\columnwidth}{!}{\input{plot}}
 \end{minipage}\hspace{2pc}%
 \begin{minipage}{18pc}
-%\includegraphics[width=18pc]{Fig4.png}
-\resizebox{\columnwidth}{!}{\input{CT2}}
+\includegraphics[width=\columnwidth]{CT174.eps}
 \end{minipage} 
-\caption{\label{fig2}Normal (left) and tangential (right) force coefficients during pitching motions of NACA0021 airfoil with a maximum amplitude of 22.6\degree\ (analog to $\lambda = 2.60$).}
+\caption{\label{fig174}Normal (left) and tangential (right) force coefficients during pitching motions of NACA0021 airfoil with a maximum amplitude of 17.4\degree\ (analog to $\lambda = 3.34$).}
 \end{figure}
 
-\newpage
+\begin{figure}[h]
+\begin{minipage}{18pc}
+\includegraphics[width=\columnwidth]{CN226.eps}
+%\resizebox{\columnwidth}{!}{\input{plot}}
+\end{minipage}\hspace{2pc}%
+\begin{minipage}{18pc}
+\includegraphics[width=\columnwidth]{CT226.eps}
+\end{minipage} 
+\caption{\label{fig226}Normal (left) and tangential (right) force coefficients during pitching motions of NACA0021 airfoil with a maximum amplitude of 22.6\degree\ (analog to $\lambda = 2.60$).}
+\end{figure}
 
-\section{Results and discussion}
+\subsection{Turbulence model comparison}
+A simulation with a RANS $k$-$\epsilon$ standard turbulence model was carried out to compare the obtained results against another one using a LES Smagorinsky turbulence model.
+Figure \ref{RANSLES} shows that for the $C_N$ values there is not considerable difference, but a small improvement has been achieved for the $C_T$ values.
+Another study cases conducted for a whole turbine simulation showed that the LES turbulence model gives better results than RANS.
 
-For a TSR of 3.34 (Figure \ref{fig1}), the maximum magnitude of the angle of attack is at
-17.4\degree\ which is in the stall region. Both \textit{CN} and \textit{CT} curves show the delay
-of the flow reattachment, which is characteristic of the dynamic stall
-phenomenon. Peaks of simulated values have close agreement with the measured
-data.
+\begin{figure}[h]
+\begin{minipage}{18pc}
+\includegraphics[width=\columnwidth]{CNRANSLES.eps}
+%\resizebox{\columnwidth}{!}{\input{plot}}
+\end{minipage}\hspace{2pc}%
+\begin{minipage}{18pc}
+\includegraphics[width=\columnwidth]{CTRANSLES.eps}
+\end{minipage} 
+\caption{\label{RANSLES}Normal (left) and tangential (right) force coefficients varying the size of the reference mesh, pithing motion with a maximum amplitude of 22.6\degree\ }
+\end{figure}
 
-For a TSR of 2.60 (Figure \ref{fig2}), the maximum amplitude of the angle of attack is 22.6\degree, which
-is related to a deeper stall region compared to a TSR of 3.34. This is shown
-because the ``loop'' of the curve is wider for the lower TSR, and moreover, the
-reattachment of the flow is further delayed.
-
-In general, the model show good agreement with experimental data, considering the complexity
-of the flows studied. This makes the model suitable for application in wind turbine
+\section{Conclusions}
+It have been demonstrated that the presented coupled model ALM with DSM, shows good agreement with experimental data, considering the complexity
+of the flows studied.
+Moreover, the model have been tested for a wide range of pithing amplitudes, therefore, with different stall conditions (not, small and deep stall), keeping a good results accuracy compared with experiments.
+The model is stable and accurate for the teste pithing blade motions.
+This makes the model suitable for application in VAWT
 simulations.
 
+\section*{Acknowledgment}
+Financial support from STandUP for Energy and Uppsala University, are gratefully acknowledged.
+
 \section*{References}
-%\begin{thebibliography}{9}
-%\bibitem{iopartnum} IOP Publishing is to grateful Mark A Caprio, Center for Theoretical Physics, Yale University, for permission to include the {\tt iopart-num} \BibTeX package (version 2.0, December 21, 2006) with  this documentation.
-Updates and new releases of {\tt iopart-num} can be found on \verb"www.ctan.org" (CTAN).
-%\end{thebibliography}
 

### 2016-06-08 paper.tex
Modifications after Goude comments

@@
 
-\subsection{Actuator Line Model (ALM)} Is a three-dimensional and unsteady aerodynamic
+\subsection{Actuator Line Model (ALM)} 
+The ALM is a three-dimensional and unsteady aerodynamic
 model developed by S{\o}rensen and Shen\cite{sorensen1999computation}, used to
@@
 \begin{align}
- & \nonumber \\ & \dfrac{\partial \rho \vec{V}}{\partial t} + \vec{V} \cdot \nabla \vec{V} = - \dfrac{1}{ \rho} \nabla p + \nu \nabla ^2 \vec{V} + \vec{f}	  \label{eqmom}  
+ & \nonumber \\ & \dfrac{\partial \vec{V}}{\partial t} + \vec{V} \cdot \nabla \vec{V} = - \dfrac{1}{ \rho} \nabla p + \nu \nabla ^2 \vec{V} + \vec{f}	  \label{eqmom}  
 \end{align} %\begin{align}
@@
 
-Lift and drag coefficients of a 2D profile are already known, and these are tabulated and combined with the blade element approach for determining 
+Static lift and drag coefficients of the 2D profile used for this study are already known from the technical technical report made by Sheldahl and Klimas \cite{sheldahl1981aerodynamic}, which is a popular database set containing values for a wide range of Reynolds numbers.
+These coefficients are tabulated and combined with the blade element approach for determining
 body forces acting on blades.
-Relative flow velocity $\vec{V}_{rel} $ and angle of attack $\alpha$ are calculated for each blade using the geometric relation between the velocity of the blade $- \omega r$, with $\omega$ representing the angular (rotational) velocity of the rotor and $r$ is the radius of the blade element, and the velocity of the local incoming wind flow $\vec{V}_{in}$.
+Relative flow velocity $\vec{V}_{rel} $ and angle of attack $\alpha$ are calculated for each blade using the geometric relation between the velocity of the blade $- \Omega r$, with $\omega$ representing the angular (rotational) velocity of the rotor and $r$ is the radius of the blade element, and the velocity of the local incoming wind flow $\vec{V}_{in}$.
 Figure \ref{figvectors} depicts a cross-sectional airfoil element at radius $r$ in the plane perpendicular to the turbine axis.
@@
 \begin{align}
-& \vec{f}_L = \frac{1}{2} \ \rho \ c \ C_L \left| \vec{V}_{rel}^2 \right|  \label{lift}  
+& f_L = \frac{1}{2} \ \rho \ c \ C_L \left| V_{rel} \right|^2  \label{lift}  
 \end{align} %\begin{align}
 \begin{align}
-& \vec{f}_D = \frac{1}{2} \ \rho \ c \ C_D \left| \vec{V}_{rel}^2 \right|  \label{drag}  
+& f_D = \frac{1}{2} \ \rho \ c \ C_D \left| V_{rel} \right|^2  \label{drag}  
 \end{align} %\begin{align}
 where $C_L$ and $C_D$ are the lift and drag coefficients, respectively.
-Both are in function of the Reynolds number and the angle of attack.
+Both are function of the Reynolds number and the angle of attack.
 It should be emphasized that the direction of the lift and drag are respectively perpendicular and parallel to the relative velocity $\vec{V}_{rel}$.
@@
 
-\subsection{Dynamic Stall Modeling (DSM)} Is represented by the Leishman-Beddoes
+\subsection{Dynamic Stall Modeling (DSM)} The DSM is represented by the Leishman-Beddoes
 model\cite{leishman1986generalised} with the modifications of Sheng et
@@
 \begin{align}
-& \Delta s| = \frac{2 | \vec{V}_{rel} | \Delta t}{c}		.	\label{deltas}
+& \Delta s = \frac{2 | \vec{V}_{rel} | \Delta t}{c}		.	\label{deltas}
 \end{align}
@@
 \end{align}
-here, $T_f$ is empirically derived time constant and its value is $T_f$ = 3. Then, the normal force coefficient for unsteady conditions before the dynamic stall onset is calculated as
+Here, $T_f$ is empirically derived time constant and its value is $T_f$ = 3. Then, the normal force coefficient for unsteady conditions before the dynamic stall onset is calculated as
 \begin{align}
@@
 Accuracy and stability were not affected even for big values for $\Delta t$, as long they are within the limit given by the CFL condition; a Courant number of less than unity is required.
-On the other hand, it should be considered that small time-step of discretization could cause numerical instabilities due to the fluctuation of the flow fields resolving the transient term $\frac{\partial}{\partial t}$.
+On the other hand, it should be considered that small time-step of discretization could cause numerical instabilities due to the fluctuation of the flow fields resolving the transient term $\frac{\partial \alpha}{\partial t}$.
 
@@
 For a TSR of 3.34 (Figure \ref{fig174}), the maximum magnitude of the angle of attack is at
-17.4\degree\ which is in the stall region. Both \textit{CN} and \textit{CT} curves show the delay
+17.4\degree\ which is in the stall region. Both $C_N$ and $C_T$ curves show the delay
 of the flow reattachment, which is characteristic of the dynamic stall
@@
 Figure \ref{RANSLES} shows that for the $C_N$ values there is not considerable difference, but a small improvement has been achieved for the $C_T$ values.
-Another study cases conducted for a whole turbine simulation showed that the LES turbulence model gives better results than RANS.
 

### 2016-06-11 paper.tex
Turn off DSM test and figures improvement with GNUPLOT

@@
 \begin{minipage}{18pc}
-\includegraphics[width=\columnwidth]{CN226revolutions.eps}
+\resizebox{\columnwidth}{!}{\input{./figures/CN226revolutions}}
 %\resizebox{\columnwidth}{!}{\input{plot}}
@@
 \begin{minipage}{18pc}
-\includegraphics[width=\columnwidth]{CT226revolutions.eps}
+\resizebox{\columnwidth}{!}{\input{./figures/CT226revolutions}}
 \end{minipage} 
@@
 \begin{minipage}{18pc}
-\includegraphics[width=\columnwidth]{CNmesh.eps}
+\resizebox{\columnwidth}{!}{\input{./figures/CNmesh}}
 %\resizebox{\columnwidth}{!}{\input{plot}}
@@
 \begin{minipage}{18pc}
-\includegraphics[width=\columnwidth]{CTmesh.eps}
+\resizebox{\columnwidth}{!}{\input{./figures/CTmesh}}
 \end{minipage} 
@@
 \begin{minipage}{18pc}
-\includegraphics[width=\columnwidth]{CN226dtLEScoarse.eps}
+\resizebox{\columnwidth}{!}{\input{./figures/CN226dtLEScoarse}}
 %\resizebox{\columnwidth}{!}{\input{plot}}
@@
 \begin{minipage}{18pc}
-\includegraphics[width=\columnwidth]{CT226dtLEScoarse.eps}
+\resizebox{\columnwidth}{!}{\input{./figures/CT226dtLEScoarse}}
 \end{minipage} 
@@
 \begin{minipage}{18pc}
-\includegraphics[width=\columnwidth]{CN138.eps}
+\resizebox{\columnwidth}{!}{\input{./figures/CN138}}
 %\resizebox{\columnwidth}{!}{\input{plot}}
@@
 \begin{minipage}{18pc}
-\includegraphics[width=\columnwidth]{CT138.eps}
+\resizebox{\columnwidth}{!}{\input{./figures/CT138}}
 \end{minipage} 
@@
 \begin{minipage}{18pc}
-\includegraphics[width=\columnwidth]{CN174.eps}
+\resizebox{\columnwidth}{!}{\input{./figures/CN174}}
 %\resizebox{\columnwidth}{!}{\input{plot}}
@@
 \begin{minipage}{18pc}
-\includegraphics[width=\columnwidth]{CT174.eps}
+\resizebox{\columnwidth}{!}{\input{./figures/CT174}}
 \end{minipage} 
@@
 \begin{minipage}{18pc}
-\includegraphics[width=\columnwidth]{CN226.eps}
+\resizebox{\columnwidth}{!}{\input{./figures/CN226}}
 %\resizebox{\columnwidth}{!}{\input{plot}}
@@
 \begin{minipage}{18pc}
-\includegraphics[width=\columnwidth]{CT226.eps}
+\resizebox{\columnwidth}{!}{\input{./figures/CT226}}
 \end{minipage} 
@@
 \begin{minipage}{18pc}
-\includegraphics[width=\columnwidth]{CNRANSLES.eps}
+\resizebox{\columnwidth}{!}{\input{./figures/CNRANSLES}}
 %\resizebox{\columnwidth}{!}{\input{plot}}
@@
 \begin{minipage}{18pc}
-\includegraphics[width=\columnwidth]{CTRANSLES.eps}
+\resizebox{\columnwidth}{!}{\input{./figures/CTRANSLES}}
 \end{minipage} 
-\caption{\label{RANSLES}Normal (left) and tangential (right) force coefficients varying the size of the reference mesh, pithing motion with a maximum amplitude of 22.6\degree\ }
+\caption{\label{RANSLES}Normal (left) and tangential (right) force coefficients for different turbulence models, pithing motion with a maximum amplitude of 22.6\degree\ }
+\end{figure}
+
+\subsection{Unsteady effects correction}
+A test for the simulation with the DSM turned off was carried out, it means there is no compensation for the unsteady effects: the effective angle of attack is the angle of attack provided by the ALM.
+Figure \ref{AMP174corrections} depicts that simulated results are closer to experimental data with the corrections.
+Both $C_N$ and $C_T$ peak values are not overestimated.
+The $C_N$ loop, shows that the model without the corrections was not able to capture reattachment flow effects.
+
+\begin{figure}[h]
+\begin{minipage}{18pc}
+\resizebox{\columnwidth}{!}{\input{./figures/CNAMP174corrections}}
+%\resizebox{\columnwidth}{!}{\input{plot}}
+\end{minipage}\hspace{2pc}%
+\begin{minipage}{18pc}
+\resizebox{\columnwidth}{!}{\input{./figures/CTAMP174corrections}}
+\end{minipage} 
+\caption{\label{AMP174corrections}Normal (left) and tangential (right) force coefficients with and without unsteady effects correction, pithing motion with a maximum amplitude of 17.4\degree\ }
 \end{figure}

### 2016-06-13 paper.tex
Wrapping and add space before citation

@@
 \section{Introduction}
-Operating conditions of Vertical Axis Wind Turbines VAWTs are characterized as complex unsteady flows which
-give a considerable challenge, both to describe using measurements and to
-represent through simulation tools\cite{huyer1996unsteady}. Moreover, VAWT
-blades are inherently exposed to cyclic variation in the angle of attack, which
-gives cyclic blade forces and can give material fatigue damage. Accurate
-modeling of the varying forces is therefore very important for the design of the
-VAWT.
+
+Operating conditions of Vertical Axis Wind Turbines VAWTs are characterized as
+complex unsteady flows which give a considerable challenge, both to describe
+using measurements and to represent through simulation
+tools~\cite{huyer1996unsteady}. Moreover, VAWT blades are inherently exposed to
+cyclic variation in the angle of attack, which gives cyclic blade forces and can
+give material fatigue damage. Accurate modeling of the varying forces is
+therefore very important for the design of the VAWT.
 

### 2016-06-13 paper.tex
General cleanup; resolves #1

@@
     this study, a coupled model is implemented with the open-source CFD
-    toolbox OpenFOAM for solving the Navier-Stokes equations, where an actuator
+    toolbox OpenFOAM for solving the Navier--Stokes equations, where an actuator
     line model and dynamic stall model are used to compute the blade loading and
@@
 
-The amplitude of the angle of attack oscillation in a fixed pitch VAWT is increasing
-with decreased tip speed
-ration (TSR), and at low TSR (common during stall regulation), the blades will
-experience dynamic stall, where the force coefficients for the blade not only
-depend on the angle of attack, but also on the rate of change of the angle of
-attack. The aim of this study is to investigate the performance of a
-Leishman-Beddoes type dynamic stall model when implemented within an actuator
-line model, for pitching motion typical to a VAWT.
+The amplitude of the angle of attack oscillation in a fixed pitch VAWT is
+increasing with decreased tip speed ration (TSR), and at low TSR (common during
+stall regulation), the blades will experience dynamic stall, where the force
+coefficients for the blade not only depend on the angle of attack, but also on
+the rate of change of the angle of attack. The aim of this study is to
+investigate the performance of a Leishman-Beddoes type dynamic stall model when
+implemented within an actuator line model, for pitching motion typical to a
+VAWT.
 
 \section{Methodology}
-For solving the governing equations of the phenomena involved, a coupled model have
-been implemented: the actuator line model is used to calculate the flow
-velocities and thereby the angle of attack. The dynamic stall model is used to
-calculate blade force coefficients, which the actuator line model needs for the
-velocity field calculations.
 
-In this work, it was chosen to validate the model
-against wind tunnel data for a pitching blade. This will put the focus on the
-force modeling part in the simulation model.
+For solving the governing equations of the phenomena involved, a coupled model
+has been implemented: the actuator line model (ALM) samples the flow velocity
+from the Navier--Stokes solver, and thereby computes each blade element's angle
+of attack and relative velocity. The dynamic stall model is used to calculate
+blade force coefficients, which the actuator line model uses to impart the body
+force back into the flow solver.
+
+In this work, it was chosen to validate the model against wind tunnel data for a
+pitching blade. This will put the focus on the force modeling part in the
+simulation model.
 
 \subsection{Actuator Line Model (ALM)}
-The ALM is a three-dimensional and unsteady aerodynamic
-model developed by S{\o}rensen and Shen\cite{sorensen1999computation}, used to
-study the flow around wind turbines. It is a combination of a solver of
-Navier-Stokes equations with a so-called actuator line technique, in which
-blades of the turbine are represented by a radial distribution of body forces
-along lines. These forces are determined using a dynamic stall model commonly
-based on empirical data. This work uses the library turbinesFoam developed by
-Bachant\cite{bachant2015simulating}. This model is based on incompressible Navier-Stokes equation
+
+The ALM is a three-dimensional and unsteady aerodynamic model developed by
+S{\o}rensen and Shen\cite{sorensen1999computation}, used to study the flow
+around wind turbines. It is a combination of a solver of the Navier--Stokes equations
+with a so-called actuator line technique, in which blades of the turbine are
+represented by a distribution of body forces along lines. These forces
+are determined using a dynamic stall model commonly based on empirical data.
+This work uses the library turbinesFoam developed by Bachant et
+al.~\cite{bachant2015simulating,Bachant2016-VAT-ALM}.
+
+The model employed here is based on the incompressible Navier--Stokes equation
 \begin{align}
@@
 
-Static lift and drag coefficients of the 2D profile used for this study are already known from the technical technical report made by Sheldahl and Klimas \cite{sheldahl1981aerodynamic}, which is a popular database set containing values for a wide range of Reynolds numbers.
-These coefficients are tabulated and combined with the blade element approach for determining
-body forces acting on blades.
-Relative flow velocity $\vec{V}_{rel} $ and angle of attack $\alpha$ are calculated for each blade using the geometric relation between the velocity of the blade $- \Omega r$, with $\omega$ representing the angular (rotational) velocity of the rotor and $r$ is the radius of the blade element, and the velocity of the local incoming wind flow $\vec{V}_{in}$.
-Figure \ref{figvectors} depicts a cross-sectional airfoil element at radius $r$ in the plane perpendicular to the turbine axis.
+Static lift and drag coefficients of the 2D profile used for this study are
+taken from the technical technical report made by Sheldahl and Klimas
+\cite{sheldahl1981aerodynamic}, which is a popular database containing values
+for a wide range of Reynolds numbers. These coefficients are tabulated and
+combined with the blade element approach for determining body forces acting on
+blades. Relative flow velocity $\vec{V}_{\mathrm{rel}}$ and angle of attack
+$\alpha$ are calculated for each blade using the geometric relation between the
+velocity of the blade $- \omega r$, with $\omega$ representing the angular
+(rotational) velocity of the rotor and $r$ the radius of the blade element, and
+the velocity of the local incoming wind flow $\vec{V}_{\mathrm{in}}$. Figure
+\ref{figvectors} depicts a cross-sectional airfoil element at radius $r$ in the
+plane perpendicular to the turbine axis.
 
@@
 \end{align} %\begin{align}
-where $C_L$ and $C_D$ are the lift and drag coefficients, respectively.
-Both are function of the Reynolds number and the angle of attack.
-It should be emphasized that the direction of the lift and drag are respectively perpendicular and parallel to the relative velocity $\vec{V}_{rel}$.
-$c$ represents the chord length.
+where $C_L$ and $C_D$ are the lift and drag coefficients, respectively. Both are
+function of the Reynolds number and the angle of attack. It should be emphasized
+that the direction of the lift and drag are respectively perpendicular and
+parallel to the relative velocity $\vec{V}_{\mathrm{rel}}$. $c$ represents the
+chord length.
 
-Same procedure is used to calculate forces from the turbine shaft and blade support struts.
-After all these forces on the actuator lines are calculated, then are added as a body force source, per unit of density (incompressible assumption), in the momentum Navier-Stokes equations (equation \ref{eqmom}).
+Same procedure is used to calculate forces from the turbine shaft and blade
+support struts. After all these forces on the actuator lines are calculated,
+then are added as a body force source, per unit of density (incompressible
+assumption), in the momentum Navier--Stokes equations (equation \ref{eqmom}).
 
-\subsection{Dynamic Stall Modeling (DSM)} The DSM is represented by the Leishman-Beddoes
-model\cite{leishman1986generalised} with the modifications of Sheng et
-al\cite{sheng2008modified} and Dyachuk\cite{dyachuk}.
-It is capable to calculate the unsteady lift,
-pitching moment and drag, giving a physical description of the aerodynamics. It
-have been validated with experimental data in\cite{leishman1989semi}.
-The model is separated into three subsystems: an attached flow model for unsteady linear
-airloads, a separated flow model for non-linear airloads and a dynamic stall
-model for the airloads induced by the leading edge vortex.
+\subsection{Dynamic Stall Modeling (DSM)} The DSM is represented by the
+Leishman-Beddoes model~\cite{leishman1986generalised} with the modifications of
+Sheng et al.~\cite{sheng2008modified} and Dyachuk~\cite{dyachuk}. It is capable to
+calculate the unsteady lift, pitching moment and drag, giving a physical
+description of the aerodynamics. It have been validated with experimental data
+in~\cite{leishman1989semi}. The model is separated into three subsystems: an
+attached flow model for unsteady linear airloads, a separated flow model for
+non-linear airloads and a dynamic stall model for the airloads induced by the
+leading edge vortex.
 
@@
 
-The model is more sensitive to the spatial discretization than the temporal one; same characteristic was obtained by Bachant et al.
-in \cite{Bachant2016-VAT-ALM}.
-Accuracy and stability were not affected even for big values for $\Delta t$, as long they are within the limit given by the CFL condition; a Courant number of less than unity is required.
-On the other hand, it should be considered that small time-step of discretization could cause numerical instabilities due to the fluctuation of the flow fields resolving the transient term $\frac{\partial \alpha}{\partial t}$.
+The model is more sensitive to the spatial discretization than the temporal one;
+same characteristic was obtained by Bachant et al. in
+\cite{Bachant2016-VAT-ALM}. Accuracy and stability were not affected even for
+big values for $\Delta t$, as long they are within the limit given by the CFL
+condition; a Courant number of less than unity is required. On the other hand,
+it should be considered that small time-step of discretization could cause
+numerical instabilities due to the fluctuation of the flow fields resolving the
+transient term $\frac{\partial \alpha}{\partial t}$.
 
@@
 
-For a TSR of 4.19 (Figure \ref{fig138}), peaks values for both $C_N$ and $C_T$ are similar to experiments.
-In this condition, the flow is attached and the blades are not in the stall region.
-Therefore, model have good agreement with experimental data using unsteady attached angle of attack calculation.
+For a TSR of 4.19 (Figure \ref{fig138}), peaks values for both $C_N$ and $C_T$
+are similar to experiments. In this condition, the flow is attached and the
+blades are not in the stall region. Therefore, model have good agreement with
+experimental data using unsteady attached angle of attack calculation.
 
-For a TSR of 3.34 (Figure \ref{fig174}), the maximum magnitude of the angle of attack is at
-17.4\degree\ which is in the stall region. Both $C_N$ and $C_T$ curves show the delay
-of the flow reattachment, which is characteristic of the dynamic stall
-phenomenon. Peaks of simulated values have close agreement with the measured
-data. The stall onset angle flow reattachment are well predicted.
+For a TSR of 3.34 (Figure \ref{fig174}), the maximum magnitude of the angle of
+attack is at 17.4\degree\ which is in the stall region. Both $C_N$ and $C_T$
+curves show the delay of the flow reattachment, which is characteristic of the
+dynamic stall phenomenon. Peaks of simulated values have close agreement with
+the measured data. The stall onset angle flow reattachment are well predicted.
 
-For a TSR of 2.60 (Figure \ref{fig226}), the maximum amplitude of the angle of attack is 22.6\degree, which
-is related to a deeper stall region compared to a TSR of 3.34. This is shown
-because the ``loop'' of the curve is wider for the lower TSR, and moreover, the
-reattachment of the flow is further delayed. Simulated peak for $C_T$ is overestimated.
+For a TSR of 2.60 (Figure \ref{fig226}), the maximum amplitude of the angle of
+attack is 22.6\degree, which is related to a deeper stall region compared to a
+TSR of 3.34. This is shown because the ``loop'' of the curve is wider for the
+lower TSR, and moreover, the reattachment of the flow is further delayed.
+Simulated peak for $C_T$ is overestimated.
 
@@
 \section{Conclusions}
-It have been demonstrated that the presented coupled model ALM with DSM, shows good agreement with experimental data, considering the complexity
-of the flows studied.
-Moreover, the model have been tested for a wide range of pithing amplitudes, therefore, with different stall conditions (not, small and deep stall), keeping a good results accuracy compared with experiments.
-The model is stable and accurate for the teste pithing blade motions.
-This makes the model suitable for application in VAWT
-simulations.
+
+It has been demonstrated that the presented coupled model ALM with DSM, shows
+good agreement with experimental data, considering the complexity of the flows
+studied. Moreover, the model has been tested for a wide range of pithing
+amplitudes, therefore, with different stall conditions (none, shallow, and deep
+stall), results agreed well compared with experiments. The model is stable and
+accurate for the tested pithing blade motions, which makes it suitable for
+application in VAT simulations.
 

### 2016-06-13 paper.tex
Cite code repo

@@
 \usepackage{amssymb} 	%for symbols
+\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,allcolors=blue]{hyperref}
 
@@
 This work uses the library turbinesFoam developed by Bachant et
-al.~\cite{bachant2015simulating,Bachant2016-VAT-ALM}.
+al.~\cite{bachant2015simulating,Bachant2016-VAT-ALM,turbinesFoam-v0.0.7}.
 

### 2016-06-13 paper.tex
Take 'ALM' out of heading
Remove acronym from heading

@@
 
-\subsection{Actuator Line Model (ALM)}
+\subsection{Actuator Line Model}
 
@@
 
-\subsection{Dynamic Stall Modeling (DSM)} The DSM is represented by the
-Leishman-Beddoes model~\cite{leishman1986generalised} with the modifications of
-Sheng et al.~\cite{sheng2008modified} and Dyachuk~\cite{dyachuk}. It is capable to
+\subsection{Dynamic Stall Modeling}
+
+The dynamic stall model (DSM) is represented by the Leishman-Beddoes
+model~\cite{leishman1986generalised} with the modifications of Sheng et
+al.~\cite{sheng2008modified} and Dyachuk~\cite{dyachuk}. It is capable to
 calculate the unsteady lift, pitching moment and drag, giving a physical

### 2016-06-13 paper.tex
'rel' in subscript should not be italics
Last modifications
Unsteady effects chapter correction,
Force distribution subchapter added
Others minimal changes VAWT and HAWT

@@
 \begin{abstract}
-    Vertical axis turbines (VATs) can be used to extract renewable energy from
-    wind or water flows. A simpler design, low cost of maintenance, and the
+    Vertical axis wind turbines (VAWT) can be used to extract renewable energy from
+    wind flows. A simpler design, low cost of maintenance, and the
     ability to accept flow from all directions perpendicular to the rotor axis
-    are some of the most important advantages over conventional horizontal axis
-    turbines (HATs). However, VATs encounter complex and unsteady fluid
+    are some of the most important advantages over conventional horizontal axis wind
+    turbines (HAWT). However, VAWT encounter complex and unsteady fluid
     dynamics, which present significant modeling challenges. One of the most
@@
     of the present study. Dynamic stall is usually used as a passive control for
-    VAT operating conditions, hence the importance of predicting its effects. In
+    VAWT operating conditions, hence the importance of predicting its effects. In
     this study, a coupled model is implemented with the open-source CFD
@@
     experimental data of pitching airfoil in similar operating conditions as an
-    H-rotor type VAT. Numerical results show reasonable agreement with
+    H-rotor type VAWT. Numerical results show reasonable agreement with
     experimental data for pitching motion.
@@
 
-Operating conditions of Vertical Axis Wind Turbines VAWTs are characterized as
+Operating conditions of Vertical Axis Wind Turbines (VAWT) are characterized as
 complex unsteady flows which give a considerable challenge, both to describe
@@
 \begin{align}
-& f_L = \frac{1}{2} \ \rho \ c \ C_L \left| V_{rel} \right|^2  \label{lift}
+& f_L = \frac{1}{2} \ \rho \ c \ C_L \left| V_{\mathrm{rel}} \right|^2  \label{lift}
 \end{align} %\begin{align}
 \begin{align}
-& f_D = \frac{1}{2} \ \rho \ c \ C_D \left| V_{rel} \right|^2  \label{drag}
+& f_D = \frac{1}{2} \ \rho \ c \ C_D \left| V_{\mathrm{rel}} \right|^2  \label{drag}
 \end{align} %\begin{align}
@@
 assumption), in the momentum Navier--Stokes equations (equation \ref{eqmom}).
+
+\section*{Calculated force distribution}
+The applied forces in the ALM need to be distributed smoothly on several mesh points in order to avoid instability due to high gradients.
+Source term forces should be projected around the element location using a three-dimensional Gaussian kernel.
+This gives the smoothing function $\eta$, which is multiplied by the computed local force on the actuator line element and then imparted on a cell with a distance $| \vec r |$ from the actuator line element quarter chord location:
+
+\begin{align*}
+\eta = \frac{1}{\epsilon ^3 \pi ^{3/2} } \exp \left[ - \left( \frac{| \vec r |}{\epsilon} \right) ^2 \right]
+\end{align*}
+
+The smoothing width parameter $\epsilon$ is chosen by the maximum value from three contributions: one relative to the 25\% of the chord length, one to the mesh size, and one to the momentum thickness due to drag force, and it is expressed as:
+
+\begin{align*}
+\epsilon = \max \left[ \frac{c}{4},4 \sqrt[\leftroot{-2}\uproot{2}3]{V_{\mathrm{cell}}},  \frac{c C_d}{2}\right]
+\end{align*}
+
+Where ${V_{\mathrm{cell}}}$ is the cell volume. More details about force projection are in \cite{Bachant2016-VAT-ALM}.
 
@@
 \subsection{Unsteady effects correction}
-A test for the simulation with the DSM turned off was carried out, it means there is no compensation for the unsteady effects: the effective angle of attack is the angle of attack provided by the ALM.
-Figure \ref{AMP174corrections} depicts that simulated results are closer to experimental data with the corrections.
+
+According to \cite{dyachuk2015simulating}, where the dynamic stall model is combined with a vortex model, it is beneficial to omit the attached flow correction equation \ref{alphaeq} and instead use $\alpha_E$ = $\alpha_n$.
+Hence, it is of interest to investigate if this correction should be included or not when the ALM is used instead of the vortex model.
+Figure \ref{AMP174corrections} depicts that simulated results with the attached flow corrections have small improvements in the accuracy and are closer to experimental data.
+Results without corrections also have good agreement.
 Both $C_N$ and $C_T$ peak values are not overestimated.
-The $C_N$ loop, shows that the model without the corrections was not able to capture reattachment flow effects.
+This same study was carried out for a bigger angle of attack range and differences are less significant.
+This has also been tested for larger angles of attack and the differences between using equation \ref{alphaeq} or not are less significant here as the separated flow modeling dominates the force calculation.
 
@@
 accurate for the tested pithing blade motions, which makes it suitable for
-application in VAT simulations.
+application in VAWT simulations.
 

### 2016-08-19 paper.tex
small correction in paragraph
Corrections based on reviewers comments (2)

@@
 Static lift and drag coefficients of the 2D profile used for this study are
-taken from the technical technical report made by Sheldahl and Klimas
+taken from the technical report made by Sheldahl and Klimas
 \cite{sheldahl1981aerodynamic}, which is a popular database containing values
@@
 $\alpha$ are calculated for each blade using the geometric relation between the
-velocity of the blade $- \omega r$, with $\omega$ representing the angular
+velocity of the blade $- \Omega r$, with $\Omega$ representing the angular
 (rotational) velocity of the rotor and $r$ the radius of the blade element, and
-the velocity of the local incoming wind flow $\vec{V}_{\mathrm{in}}$. Figure
+the velocity of the local incident wind flow $\vec{V}_{\mathrm{in}}$ which is generally lower than the incoming flow velocity $\vec{V}_{\infty}$. Figure
 \ref{figvectors} depicts a cross-sectional airfoil element at radius $r$ in the
@@
 \begin{center}
-\includegraphics[width=0.35\columnwidth]{vector.eps}
+\includegraphics[width=0.33\columnwidth]{vector2.eps}
 \end{center}
@@
 \begin{align*}
-\epsilon = \max \left[ \frac{c}{4},4 \sqrt[\leftroot{-2}\uproot{2}3]{V_{\mathrm{cell}}},  \frac{c C_d}{2}\right]
+\epsilon = \max \left[ \frac{c}{4},4 \sqrt[\leftroot{-2}\uproot{2}3]{V_{\mathrm{cell}}},  \frac{c C_D}{2}\right]
 \end{align*}
@@
 \end{align}
-where $\alpha$ is the geometrical angle of attackm $X$, $Y$ and $Z$ are the deficiency functions, which are empirically derived, based on the flow velocity and the pithing rate, the details are found in \cite{dyachuk2013dynamic}.
-Indices $n$ and $n-1$ in equations \ref{CNcirc} and \ref{alphaeq} represent the current and previous time-step.
+where $\alpha$ is the geometrical angle of attack and $X$, $Y$ and $Z$ are the deficiency functions, which are empirically derived, based on the flow velocity and the pithing rate, the details are found in \cite{dyachuk2013dynamic}.
+Indices $n$ and $n-1$ represent the current and previous time-step.
 A delayed angle of attack is considered due to the lag in pressure response, and it is calculated as
@@
 \begin{align}
-& C_{N_n}^f = C_{N_\alpha} \alpha_{E_n} \left( \frac{1 + \sqrt{f''}}{2} \right)^2	\label{CNf}
+& C_{N_n}^f = C_{N_\alpha} \alpha_{E_n} \left( \frac{1 + \sqrt{f''_n}}{2} \right)^2	\label{CNf}
 \end{align}
@@
 \begin{align}
-C_N^v = B_1 (f'' - f) V_x \label{CNv}
+C_N^v = B_1 (f''_n - f_n) V_x \label{CNv}
 \end{align}
@@
 %\end{minipage}%
-\caption{\label{figpitching} Illustration of dynamic stall: NACA0021 profile with $\alpha=6 + 16 \sin (\omega	t), \ \omega=12,47$ [rad/s], $V_\infty=34,3$ [m/s] and $c$ = 0,55[m]}
+\caption{\label{figpitching} Illustration of dynamic stall: NACA0021 profile with $\alpha=6 + 16 \sin (\Omega	t), \ \Omega=12,47$[rad/s], $V_\infty=34,3$[m/s] and $c$ = 0,55[m]}
 \end{figure}
@@
 \begin{align}
-C_T = \eta C_{N_\alpha} \alpha_E ^2 ( \sqrt{f''}-E_0)     \label{CT}
+C_T = \eta C_{N_\alpha} \alpha_E ^2 ( \sqrt{f''_n}-E_0)     \label{CT}
 \end{align}
@@
 In the equation \ref{alphaval}, $\alpha$ represents the geometric angle of attack, and it is different to the effective angle of attack $\alpha_E$, which is calculated by the model in order to get the force coefficients.
-A NACA0021 profile with 0.55 [m] of chord length is driven to the dynamic stall region trough the pitching motion accordint to equation \ref{alphaval}.
+A NACA0021 profile with 0.55[m] of chord length is driven to the dynamic stall region trough the pitching motion accordint to equation \ref{alphaval}.
 The pitching blade experiments were carried out at Glasgow
@@
 Varying parameters have been tested, mainly in the deep stall condition for an amplitude of 22.6 $\degree $ (equivalent to $\lambda$ = 2.6).
-All the studied cases were simulated using a LES Smagorinsky turbulence model.
+All the studied cases were simulated with a domain geometrically similar to the wind tunnel used in \cite{angell1988collected} to emulate the pitching blade experiments: 1.61[m] of span, 2.13[m] of height, 2.5[m] upstream and 4[m] downstream, with similar velocity boundary conditions: 28.4[m/s] in the inlet and no-slip condition (fixed 0[m/s]) in the walls, using a LES Smagorinsky turbulence model.
 Figure \ref{revolutions} shows that the model needs two revolutions to reach the convergence, a third revolution will not produce any change.
@@
 
-\textbf{Temporal sensitivity:} Different time-steps were chosen for a varying temporal discretization test, which $\Delta t$ values are 0.001, 0.003, 0.0055 and 0.0075 [s], and these correspond to around 1200, 400, 220 and 160 time-steps per revolution respectively.
+\textbf{Temporal sensitivity:} Different time-steps were chosen for a varying temporal discretization test, which $\Delta t$ values are 0.001, 0.003, 0.0055 and 0.0075[s], and these correspond to around 1200, 400, 220 and 160 time-steps per revolution respectively.
 In this study a coarse mesh with a refinement of 25\%, compared to the reference one, was used.
@@
 Both $C_N$ and $C_T$ peak values are not overestimated.
-This same study was carried out for a bigger angle of attack range and differences are less significant.
 This has also been tested for larger angles of attack and the differences between using equation \ref{alphaeq} or not are less significant here as the separated flow modeling dominates the force calculation.

### 2016-08-21 paper.tex
Pithing --> pitching

@@
 \end{minipage}
-\caption{\label{RANSLES}Normal (left) and tangential (right) force coefficients for different turbulence models, pithing motion with a maximum amplitude of 22.6\degree\ }
+\caption{\label{RANSLES}Normal (left) and tangential (right) force coefficients for different turbulence models, pitching motion with a maximum amplitude of 22.6\degree\ }
 \end{figure}
@@
 \end{minipage}
-\caption{\label{AMP174corrections}Normal (left) and tangential (right) force coefficients with and without unsteady effects correction, pithing motion with a maximum amplitude of 17.4\degree\ }
+\caption{\label{AMP174corrections}Normal (left) and tangential (right) force coefficients with and without unsteady effects correction, pitching motion with a maximum amplitude of 17.4\degree\ }
 \end{figure}

### 2016-08-21 paper.tex
Made by --> by
More instances of 'pithing'

@@
 Static lift and drag coefficients of the 2D profile used for this study are
-taken from the technical report made by Sheldahl and Klimas
+taken from the technical report by Sheldahl and Klimas
 \cite{sheldahl1981aerodynamic}, which is a popular database containing values
@@
 \end{align}
-where $\alpha$ is the geometrical angle of attack and $X$, $Y$ and $Z$ are the deficiency functions, which are empirically derived, based on the flow velocity and the pithing rate, the details are found in \cite{dyachuk2013dynamic}.
+where $\alpha$ is the geometrical angle of attack and $X$, $Y$ and $Z$ are the deficiency functions, which are empirically derived, based on the flow velocity and the pitching rate, the details are found in \cite{dyachuk2013dynamic}.
 Indices $n$ and $n-1$ represent the current and previous time-step.
@@
 \end{minipage}
-\caption{\label{figdtcoarse}Normal (left) and tangential (right) force coefficients varying time-step discretization for a pithing motion with a maximum amplitude of 22.6\degree\ }
+\caption{\label{figdtcoarse}Normal (left) and tangential (right) force coefficients varying time-step discretization for a pitching motion with a maximum amplitude of 22.6\degree\ }
 \end{figure}
@@
 good agreement with experimental data, considering the complexity of the flows
-studied. Moreover, the model has been tested for a wide range of pithing
+studied. Moreover, the model has been tested for a wide range of pitching
 amplitudes, therefore, with different stall conditions (none, shallow, and deep
 stall), results agreed well compared with experiments. The model is stable and
-accurate for the tested pithing blade motions, which makes it suitable for
+accurate for the tested pitching blade motions, which makes it suitable for
 application in VAWT simulations.
